The Design and Implementation of the Synchronous Language \CEU

### 2016-04-05 paper.tex
...

\author{
Francisco Sant'Anna
\affil{Departamento de Inform\'atica, PUC--Rio}
Roberto Ierusalimschy
\affil{Departamento de Inform\'atica, PUC--Rio}
Noemi Rodriguez
\affil{Departamento de Inform\'atica, PUC--Rio}
Silvana Rossetto
\affil{Departamento de Ci\^encia da Computa\c{c}\~ao, UFRJ}
Adriano Branco
\affil{Departamento de Inform\'atica, PUC--Rio}
}

@@

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with simpler semantics and 
more fine-grained control over program execution.
%
\CEU adopts an event-triggered notion of time which enables compile-time 
analysis for simultaneous reactions to produce deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we discuss the particularities of our design, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis such as to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
Both problems have an extensive coverage and debate in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
which, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of the loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to interact with stateful system calls safely and makes shared-memory 
concurrency not as straightforward as reading and writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
% TODO: simple simple
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
\CEU \bx for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 4--11), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
%We finish with a summary of our design
    %(\ref{sec.ceu.discussion}).
%
We provide the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

>>>>

\author{
Francisco Sant'Anna
\affil{Departamento de Inform\'atica e Ci\^encia da Computa\c{c}\~ao, UERJ}
Roberto Ierusalimschy
\affil{Departamento de Inform\'atica, PUC--Rio}
Noemi Rodriguez
\affil{Departamento de Inform\'atica, PUC--Rio}
Silvana Rossetto
\affil{Departamento de Ci\^encia da Computa\c{c}\~ao, UFRJ}
Adriano Branco
\affil{Departamento de Inform\'atica, PUC--Rio}
}

@@

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with simpler semantics and 
more fine-grained control over program execution.
%
\CEU adopts an event-triggered notion of time that enables compile-time 
temporal analysis to detect conflicting statements, resulting in deterministic 
and concurrency-safe programs.
%
Using Esterel as a base, we discuss the particularities of our design such as 
stack-based internal events, the temporal analysis, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis such as to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
which, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of loose and 
non-deterministic execution for intra-reaction statements.
Non determinism prevents threads to interact with stateful system calls safely 
and makes shared-memory concurrency not as straightforward as reading and 
writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
% TODO: simple simple
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution, and is amenable to a temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

Figure~\ref{lst.abro} illustrates the syntactic similarities between the 
languages, showing side-by-side the implementations in Esterel \ax and \CEU \bx 
for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 4--11) which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish with a summary of our design
    (\ref{sec.ceu.discussion}).
%
We provide the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.
